Project 2

### 2015-04-25 project2/paper/main.tex
Final cleanup!

Because of this, we can represent the entire metro as queuing network with sections between stations as the queue. In
our case, we will only be modeling the section of track between Rosslyn and Stadium-Armory where the Orange, Blue, and
Silver share a track (See \ref{fig:metromap}). Since there are no track crossovers, we can consider these three lines
to be isolated from other lines. However, we do need to keep in mind that for two tracks, each track is devoted to a
specific direction. Only in the case of a single track is when a track will handle both directions.

@@

Because this will be a discrete-event simulation, we will need to be mindful of the events that travel through our
application. A first example that comes to mind is the \texttt{train:arrive} event, which will be received by the
Rosslyn and Stadium-Armory stations when a train first arrives in the system. This event will need to contain the train
instance, along with the direction its heading (although that is implied by the station in our case). This way, we can
track the train as it moves through the system to calculate metrics about the system's performance.

@@

\texttt{train:exit} is triggered when a train exits a connection, which informs a connection to free up the track and
push the next train through if there is one. As mentioned before, this event is actually scheduled when a train enters
a track between stations. Each connection has knowledge of the time between it's associated stations, and uses that
time to schedule the \texttt{train:exit} in relation to \texttt{train:enter} for a particular train. Lastly, when a
train exits a particular connection, a \texttt{train:arrive} event is scheduled immediately for the next connection in
the metro system.

@@

\subsection{Query Design}
Viewing this as a queueing problem, there are some ``missing pieces'' that our model did not incorporate, namely the
arrival rate of trains and the service time for a single train. Instead of attempting to calculate these based on 
numerical models of motion, we can instead use the WMATA API \cite{wmataapi} to calculate this.

@@

Given this information, we can calculate how often a train runs simply by marking the timestamp of the API call and
offset by the estimated time until arrival provided by the API. For instance, if we query a station at 12:00 PM, and
the API estimates a train arriving in 3 minutes, we can roughly estimate that the train will arrive at 12:03. This
gives us our arrival rate. Furthermore, we can query by train line as well to verify that trains run according to the
schedule that WMATA provides on its website \cite{metroschedule}.

@@

\subsection{Query Results}
A script was built to call the WMATA API for train estimates, and was setup as a cron job on server. The script
essentially downloaded the estimates of train arrivals for every station into a single timestamped JSON file (see the
JSON format \cite{json}. Cron \cite{crontab} was configured to run the script every minute for an entire week,
resulting in a directory full of timestamped JSON files. These were then imported into a Mongo \cite{mongodb} database 
to allow easier queries.

Train interarrivals were measured by line and direction. \ref{fig:eastboundblueinterarrivals} demonstrates a case where
the data appears to be normal. However, other cases such as \ref{fig:westboundsilverinterarrivals} show that this is
not necessarily true. Graphs for the remaining directions and line colors were omitted for brevity, but are included in
the submission. Due to time constraints, we'll use an empirical distribution rather than trying to fit any particular
distribution.

@@

One of the simplifications that we've made is that we've excluded all other stations in the metro system except those
between Rosslyn and Stadium Armory. Other lines, the Red, Green, and Yellow lines, do not share track with any train,
nor do they cross at any point. In this respect, they are isolated (regardless of the fact that they may share
stations). So, in this case, the simplification hinders accuracy in no way whatsoever. However, ignoring Blue, Orange,
and Silver stations beyond Stadium-Armory and Rosslyn could potentially
reduce model accuracy here because train events at those stations can very easily influence the stations within D.C.
We have chosen instead to model those exterior stations as a probability distribution which is based off of empirical
data, but we still run a risk that we may have not captured an important interaction.

@@

\section{Technology Used}
The simulation was built using CoffeeScript \cite{coffeescript}, using Backbone \cite{backbone}, Underscore
\cite{underscore} and jQuery \cite{jQuery} to organize code some. Code was bundled via Browserify \cite{browserify},
and minified with UglifyJS \cite{uglify}. Gulp \cite{gulp} was used as a build tool and package management was done
via NPM \cite{npm}. Geo-mapping was done via Leaflet \cite{leaflet}, which provides an API for enhancing the map with
markers, text, shapes, etc.

In an effort to ensure code is working properly, the simulation was built with a testing suite. Mocha \cite{mocha} was
used as the testing framework, and Chai \cite{chai} was used for BDD-style assertions. Sinon \cite{sinon} was used to
mock function calls during tests.

@@

Regardless of these discrepancies, the simulation appears to agree with the estimates. Trains moving westward average
around 22.15 minutes to reach Rosslyn from Stadium-Armory. Eastward trains perform slightly better, hovering closer to
21 minutes. Both distributions, without any track disturbances, are very tight with little variance. The difference in
means is accounted for by the empirical measurement of between station times. The total westward time from Rosslyn to
Stadium-Armory is roughly 30 seconds shorter than the eastward counterpart. The remaining time may be accounted for by
the Rosslyn-Foggy Bottom connection, the most western connection in the system. The between station time here is 
consistently 3 minutes, rather than 2. Because of this, arriving trains may be forced to wait more often. However,
because our scope is narrowed for this project, we may be missing some other behavior further westward that may explain
this behavior.

Unsurprisingly, WMATA plans its trains effectively to be robust to a fair amount track outages, while still providing
the timings estimated on their website. Even with three tracks disabled, the mean travel time of eastern and western
trains still hovers around 23 minutes and 30 seconds. However, worth noting is that the \emph{variabilty} increases
drastically. Even though the mean moves little, there still are some trains that end up taking over 40 minutes to
reach the other side given these conditions. In a way, this is simply ``bad luck'' on that particular train's part,
simply because it's timing fell so that it arrived right as another train had occupied the track (with potentially
other trains waiting already). However as more tracks become disabled, the bad luck slowly becomes more likely for any
train to experience.

>>>>

Because of this, we can represent the entire metro as queuing network with sections between stations as the queue. In
our case, we will only be modeling the section of track between Rosslyn and Stadium-Armory where the Orange, Blue, and
Silver share a track (See \ref{fig:metromap}) which results in a simpler network with the nodes falling in a single
line. Since there are no track crossovers within the DC metro, we can consider these three lines
to be isolated from other lines, so we lose no accuracy by excluding the other lines from the simulation. However, we
do need to keep in mind that for two tracks, each track is devoted to a specific direction. Only in the case of a single
track is when a track will handle both directions.

@@

Because this will be a discrete-event simulation, we will need to be mindful of the events that travel through our
application. A first example that comes to mind is the \texttt{train:arrive} event, which will be received by the
Rosslyn and Stadium-Armory stations when a train first arrives in the system. This event will need to contain the train
instance, along with the direction its heading. This way, we can
track the train as it moves through the system to calculate metrics about the system's performance.

@@

\texttt{train:exit} is triggered when a train exits a connection, which informs a connection to free up the track and
push the next train through, if there is one. As mentioned before, this event is actually scheduled when a train enters
a track between stations. Each connection has knowledge of the time between it's associated stations, and uses that
time to schedule the \texttt{train:exit} in relation to \texttt{train:enter} for a particular train. Lastly, when a
train exits a particular connection, a \texttt{train:arrive} event is scheduled immediately for the next connection in
the metro system.

@@

\subsection{Query Design}
Viewing this as a queueing problem, there are some ``missing pieces'' that our model did not incorporate, namely the
arrival rate of trains and the service time for a single train. Instead of attempting to calculate these based on 
numerical models of motion, we can instead use the WMATA API \cite{wmataapi} to claculate an empirical value.

@@

Given this information, we can calculate how often a train runs simply by marking the timestamp of the API call and
offset by the estimated time until arrival provided by the API. For instance, if we query a station at 12:00 PM, and
the API estimates a train arriving in 3 minutes, we can roughly estimate that the train will arrive at 12:03.
Performing this operation over time gives us a list of train arrival time by station from which we can calculate
interarrival time via a running diff. Furthermore, we can query by train line as well to verify that trains run
according to the schedule that WMATA provides on its website \cite{metroschedule}.

@@

\subsection{Query Results}
A script was built to call the WMATA API for train estimates, and was setup as a cron job on a server and scheduled to
run every minute for a week. The script
essentially downloaded the estimates of train arrivals for every station into a single timestamped JSON file (see the
JSON format \cite{json}. Cron \cite{crontab} was configured to run the script every minute for an entire week,
resulting in a directory full of timestamped JSON files. These were then imported into a MongoDB \cite{mongodb} database 
to allow easier queries.

Train interarrivals were measured by line and direction. (\ref{fig:eastboundblueinterarrivals}) demonstrates that the
data oftentimes appears to follow some ``standard'' distribution. However, other cases such as
(\ref{fig:westboundsilverinterarrivals}) show that this is not necessarily true. Graphs for the remaining directions
and line colors were omitted for brevity, but are included in
the submission. Due to time constraints, we'll use an empirical distribution rather than trying to fit any particular
distribution.

@@

One of the simplifications that we've made is that we've excluded all other stations in the metro system except those
between Rosslyn and Stadium Armory. Other lines, the Red, Green, and Yellow lines, do not share track with any train,
nor do they cross at any point. In this respect, they are isolated. Event though other lines may share stations, the
trains run at a different elevation and therefore do no share a track. So, in this case, the simplification hinders
accuracy in no way whatsoever. However, ignoring Blue, Orange, and Silver stations beyond Stadium-Armory and Rosslyn
could potentially
reduce model accuracy here because train events at those stations can very easily influence the stations within D.C.
We have chosen instead to model those exterior stations as a probability distribution which is based off of empirical
data, but we still run a risk that we may have not captured an important interaction.

@@

\section{Technology Used}
The simulation was built using CoffeeScript \cite{coffeescript}, Backbone \cite{backbone}, Underscore
\cite{underscore} and jQuery \cite{jQuery} to organize code some. Code was bundled via Browserify \cite{browserify},
and minified with UglifyJS \cite{uglify}. Gulp \cite{gulp} was used as a build tool and NPM \cite{npm} was used as a
package manager. Geo-mapping was provided by Leaflet \cite{leaflet}, which has an API for enhancing the map with
markers, text, shapes, etc.

In an effort to ensure that the code worked properly, the simulation was built with a testing suite. Mocha \cite{mocha}
was used as the testing framework, and Chai \cite{chai} was used for BDD-style assertions. Sinon \cite{sinon} was used
to mock function calls during tests.

@@

Regardless of these discrepancies, the simulation appears to agree with the estimates. Trains moving westward average
around 22.15 minutes to reach Rosslyn from Stadium-Armory within the simulation. Eastward trains perform slightly
better, hovering closer to 21 minutes. Both distributions, without any track disturbances, are very tight with little
variance. The difference in  means is accounted for by the empirical measurement of between station times. The total
westward time from Rosslyn to  Stadium-Armory is roughly 30 seconds shorter than the eastward counterpart. This could
be a result of many factors, such as geography, or even interactions beyond Rosslyn and Stadium-Armory ``captured''
in our empirical distribution. The remaining time may be accounted for by the Rosslyn-Foggy Bottom connection, the
most western connection in the system. The between station time here is  consistently 3 minutes, rather than 2.
Because of this, arriving trains may be forced to wait more often. WMATA estimates the travel time being the same
regardless of direction, and I still cannot explain the discrepancy.

Unsurprisingly, WMATA plans its trains effectively to be robust to a fair amount track outages, while still providing
the timings estimated on their website. Even with three tracks disabled, the mean travel time of eastern and western
trains still hovers around 23 minutes and 30 seconds. However, worth noting is that the \emph{variabilty} increases
drastically. Even though the mean moves little, there still are some trains that end up taking over 40 minutes to
reach the other side given these conditions. In a way, this is simply ``bad luck'' on that particular train's part,
simply because it's timing fell so that it arrived right as another train had occupied the track (with potentially
other trains waiting already). However as more tracks become disabled, the bad luck slowly becomes more likely for any
train to experience. Eventually, the mean does increase to around 32 minutes, with a very wide distribution with many
trains exceeding 45 minute travel times.

In conclusion, we see that WMATA does indeed provide fairly accurate estimates of train travel times, along with the
interarrival times between stations. Furthermore, WMATA's train scheduling is fairly robust to track outages, without
drastically impacting the throughput of trains. There are some unanswered questions, however. Why does the direction
affect the travel time between stations? Why do the simulated travel times agree with WMATA's estimate, but the average
interarrivals measured from the API do not? Would increasing the scope of the model improve the accuracy and remove the
discrepancies altogether? These questions can perhaps be answered with deeper investigation and more data.
